# Data Visualizations {.unnumbered}

### `ggplot2` components

A `ggplot` is comprised of 1. data, 2. aesthetic mappings, and 3. geometries (the type of plot you would like to graph).

Beyond the bar charts and line graphs discussed in the previous section, ggplot2 offers a variety of other visualizations that you can create. We'll be using the `mtcars` dataset, which is readily available in your R environment, to explore these options.

```{r}
library(ggplot2)
data(mtcars)
head(mtcars)
```

## Types of Plots

### Scatter Plot

-   Shows relationship between two continuous variables

```{r}
scatter <- ggplot(data = mtcars, 
                  aes(x = wt, y = mpg))+
  geom_point()+
  labs(x = "Weight",
       y = "Miles per Gallon")
print(scatter)
```

### Line Plot

-   Displays trends over time or ordered data

```{r}
line <- ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_line() +
  labs(x = "Horsepower",
       y = "Miles Per Gallon")
print(line)

# if you want to plot the trend of the different
# values in a categorical variable, you can 
# set it as a color 
line <- ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_line(aes(color = factor(gear))) +
  labs(x = "Horsepower",
       y = "Miles Per Gallon", 
       color = "Number of Gear")
print(line)
```

### Bar Plot

-   Displays counts or summary statistics of categorical data

```{r}
bar <- ggplot(mtcars, aes(x = factor(cyl))) +
  geom_bar(fill = "steelblue") +
  labs(x = "Number of Cylinders",
       y = "Count")
print(bar)
```

### Histogram

-   Displays the distribution of a single continuous variable

```{r}
hist <- ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(binwidth = 2, fill = "steelblue") +
  labs(title = "Histogram of MPG",
       x = "Miles Per Gallon",
       y = "Frequency")
print(hist)
```

### Boxplot

-   Displays the distribution of one or multiple continuous variables

```{r}
# you can fill/color the boxes to correspond to a # specific categorical variable as well 
box <- ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
  geom_boxplot(aes(fill = factor(cyl))) +
  labs(title = "Box Plot of MPG by Cylinder",
       x = "Number of Cylinders",
       y = "Miles Per Gallon", 
       fill = "Number of Cylinders")
print(box)
```

### Density Plot

-   Displays the distribution as a smoothed curve of a continuous variable
- `alpha:` controls the transparency of the `fill` color.

```{r}
density <- ggplot(mtcars, aes(x = mpg)) +
  geom_density(fill = "blue", alpha = 0.5) +
  labs(x = "Miles Per Gallon",
       y = "Density")
print(density)

```

### Violin Plot

-   To show the distribution of a continuous variable by categories (the `factor` variable), combining aspects of a box plot and a density plot.

```{r}
violin <- ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
  geom_violin(aes(fill = factor(cyl))) +
  labs(x = "Number of Cylinders",
       y = "Miles Per Gallon", 
       fill = "Number of Cylinders")
print(violin)
```

### Faceted Plots

-   Creates multiple plots based on the values of a categorical variable

```{r}
facet <- ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(aes(color = factor(cyl))) +
  facet_wrap(~ cyl) +
  labs(x = "Weight",
       y = "Miles Per Gallon (mpg)", 
       color = "Number of Cylinders")
print(facet)
```

## Themes

`ggplot2`offers a variety of built-in themes that you can apply to change the appearance of your plots. Here are a few options you can use:

-   `theme_grey()`

-   `theme_bw()`

-   `theme_linedraw()`

-   `theme_light()`

-   `theme_dark()`

-   `theme_minimal()`

-   `theme_classic()`

-   `theme_void()`

Here is how you can implement:

```{r}
# re-using the box plot, we can append a new
# ggplot element to our variable box

box <- box + theme_minimal()
print(box)
```

## Additional Customization

### Case Study: Airquality Dataset

To discuss more on how to implement further customization of your visualizations with `ggplot2`, we will be using the built-in R dataset `airquality`.

```{r}
library(tidyverse)
air <- airquality
summary(air)
```

The summary of our `air` dataset shows that it has 6 different variables, 4 continuous and 2 representing time. Let's set labels and factor our `Month` column so that it represents the named Month instead of the numeric Month. With `factor()`, we can set labels onto our levels so when defining our labels, it will automatically map to the order of its current level.

```{r}
air$Month <- factor(air$Month, labels = c("May", "June", "July", "August",
                               "September"))
```

We will be creating three distinct plots, encoding the 'Month' variable through the use of color. To visually represent time, we'll employ a gradient of blue shades, with lighter blues indicating earlier months and darker blues representing later months. We will also ensure that all 3 plots utilize the same color gradient.

```{r}
# create a vector that establishes our colors for each month 
months <- c("May", "June", "July", "August", "September")
cols   <- c("#CAF0F8", "#90E0EF", "#00B4D8",
          "#0077B6", "#03045E")

names(cols) <- months
```

Our first plot, we will visualize the distribution of `Temp` by `Month`. This will help us visualize the differences in temperature by month.

```{r}
g1 <- ggplot(data = air, aes(x=Temp, fill=Month))+
  geom_density(alpha = 0.7)+
  scale_fill_manual(name = "Months", values=cols)+
  theme(axis.text.x = element_text(size = 14), 
        axis.text.y = element_text(size = 14), 
        axis.title.x = element_text(size = 20), 
        axis.title.y = element_text(size = 20))
print(g1)
```

-   The `scale_fill_manual()` element of our ggplot ensures that the Months are colored by what we defined in the `cols` vector.

-   Within our `theme()` element, we are increasing the font size for the x-axis ticks (`axis.text.x`), x-axis title (`axis.title.x`), y-axis ticks (`axis.text.y`), and y-axis title (`axis.title.y`).

In our second plot, we will visualize the temperature trends by day. Additionally, we'll incorporate wind speed as another variable. Given that wind speed is a continuous variable, we will represent it by varying the size of our data points—smaller points for lower wind speeds and larger points for higher speeds.

```{r}
g2 <- ggplot(data = air, aes(x = 1:nrow(air), y=Temp))+
  geom_line(aes(col = Month))+
  geom_point(aes(col = Month, size = Wind))+
  geom_smooth(method = "loess", col="black")+
  scale_color_manual(name = "Months", values=cols)+
  geom_vline(xintercept=c(31, 61, 92, 123, 153), linetype="dashed")+guides(color="none")+
  labs(x="Day")
print(g2)
```

For this plot above, x is defined as `1:nrow(air)`-- the column `Day` in the `air` dataset corresponds to each month, however, since our dataset is ordered, the indices represent the true number of days.

-   The `geom_point()` element contains the argument `size` where we defined it to be `Wind`. ggplot will automatically determine what breakpoints are good to split these values/sizes up in.

-   The `scale_color_manual()` function ensures month is represented in the colors we discussed previously. Here, we used `scale_color_manual` instead of `scale_fill_manual` since our aesthetic for the points and line is a `color` rather than a fill.

-   `geom_vline()` allows you to add in vertical lines throughout the plot when provided an x-intercept. In our case, 31,61,92,123, and 153 represent the indices of our last days of the month.

-   `guides(color="none"`) removes color from our legend, which will be useful later.

Our third plot will represent the relationship between `Wind` and `Temp`, but also including `Ozone` into the plot by using `size` to represent its value. However, in this plot, we would also like to label datapoints that represent the max temperature for each month.

Given that we're incorporating Ozone into this plot, I will eliminate any NAs from the dataframe. Initially, I considered recoding these as 0, but since the Ozone column represents the average ozone in parts per billion, a zero value could mistakenly indicate an actual measurement.

```{r}
air_remove <- air %>% filter(Ozone != (is.na(Ozone)))
```

Next, we are going to find the max temperatures by each year.

```{r}
max_temp <- air %>% group_by(Month) %>%
  filter(Temp == max(Temp))
```

```{r}
g3 <- ggplot(data = air, aes(x=Wind, y=Temp))+
  geom_point(aes(size=Ozone, color=Month))+
  geom_smooth(method="lm", alpha = 0.2, aes(color = Month))+
  geom_text(data = max_temp, aes(label = Temp), vjust=-1, hjust=0.5)+
  scale_color_manual(name = "Months", values=cols)+
  guides(color="none")+theme_minimal()
print(g3)
```

-   The `geom_text()` element enables us to annotate the plot with maximum temperatures. We can create a dedicated dataset for it to process. By supplying only the `max_temp` dataframe, it will exclusively display the values contained within that dataset.